---
title: "Rmardown"
author: "Jason"
date: "`r Sys.Date()`"
output: html_document
---

### Load package

```{r}
library(tidyverse)

```

### Read file

```{r}
# create raw file directory
if(!file.exists("raw")) {
    dir.create("raw")
}

# download file
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
              destfile = "raw/data.zip",
              method = "curl")

# unzip the file
unzip("raw/data.zip", exdir = "raw")

# load the dataset
data <- read.csv("raw/activity.csv")
```

### Format data

```{r}
data$date <- ymd(data$date)
```

### What is mean total number of steps taken per day?

#### 1. Calculate the total number of steps taken per day

```{r}

no_steps_day <- data |> 
    group_by(date) |> 
    summarise(total_steps = sum(steps, na.rm = TRUE))

no_steps_day
    
```

#### 2. Total number steps taken each day in histogram

```{r}

ggplot(no_steps_day, aes(date, total_steps)) +
    geom_col() +
    labs(y = "total steps", title = "Total number of steps taken each day")

```

#### 3. Median and mean of the total number of steps taken per day

```{r}

mean <- mean(no_steps_day$total_steps)
median <- median(no_steps_day$total_steps)

```

The mean is `r mean` The median is `r median`

### What is the average daily acitivity pattern?

#### 1. Time-series plot

```{r}

# format the interval col in hour-minute format
no_steps_interval <- data |> 
    group_by(interval) |> 
    summarise(average_steps = mean(steps, na.rm = TRUE))

# plot the time series graph
ggplot(no_steps_interval, aes(interval, average_steps)) +
    geom_line() +
    labs(y = "average steps", title = "Average steps per interval")

```

#### 2. Which interval has the highest step

```{r}
index <- which(no_steps_interval$average_steps == max(no_steps_interval$average_steps, na.rm = TRUE))

interval <- no_steps_interval[index, "interval"]
```

Time `R interval` has the highest number of steps

### Inputting missing values

#### 1. Calculate and report the total number of NAs in steps col

```{r}
missing_data <- data[is.na(data$steps), "steps"]
length_missing <- length(missing_data)
```

There are `R length_missing` missing values in the data

#### 2. Fill in the missing values with the mean of that interval

```{r}
data_missing_filled <- data |> 
    left_join(no_steps_interval, 
              by = "interval") |> 
    mutate(steps = ifelse(is.na(steps),
                           average_steps,
                           steps))
```

#### 3. New df with NAs filled in

```{r}
data_missing_filled
```

#### 4. Histogram of total steps per day, mean and median

```{r}

no_steps_day_missing_filled <- data_missing_filled |> 
    group_by(date) |> 
    summarise(total_steps = sum(steps))

ggplot(no_steps_day_missing_filled, aes(date, total_steps)) +
    geom_col()

mean_missing_filled <- mean(no_steps_day_missing_filled$total_steps)

median_missing_filled <- median(no_steps_day_missing_filled$total_steps)
```

The updated mean and median after the NAs are filled with the daily average are `R mean_missing_filled` and `R median_missing_filled`

Filling the NAs with the daily average has increased the mean and median slightly

### Difference in activity patterns between weekend and weekdays

#### 1. Create a new var indicating weekend or weekday

```{r}
data_missing_filled_wday <- data_missing_filled |> 
    mutate(weekday = weekdays(data_missing_filled$date,
                              abbreviate = FALSE))

data_missing_filled_wday <- data_missing_filled_wday |> 
    mutate(dayOrend = ifelse(weekday %in% c("Saturday", "Sunday"),
                             "weekend",
                             "weekday"))

data_missing_filled_wday
```

#### 2. Panel plot of avrage steps taken per interval between weekend and weekday

```{r}
no_steps_interval_NAfilled <- data_missing_filled_wday |> 
    group_by(interval, dayOrend) |> 
    summarise(average_steps = mean(steps))

ggplot(no_steps_interval_NAfilled, aes(interval, average_steps)) +
    geom_line() +
    labs(y = "average steps", title = "Average steps taken by interval: weekday vs weekend") +
    facet_wrap(~ dayOrend)
    
```
